The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-13 talk/vmil2012/paper.tex
improvements for background section

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations.
\bivab{Mention the different language impls}

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as (if needed) a garbage collector
and a just-in-time compiler based on hints provided by the author.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing is a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs typically are
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

@@

Figure~\ref{fig:example} shows an example RPython function that checks
whether a number reduces to 1 with less than 100 steps of the Collatz process.
It uses an \lstinline{Even} and an \lstinline{Odd} class to box the numbers, to
make the example more interesting. If the loop in \lstinline{check_reduces} is
traced when \lstinline{a} is a multiple of four, the unoptimized
trace looks like in Figure~\ref{fig:unopt-trace}. The line numbers in the trace
correspond to the line numbers in Figure~\ref{fig:trace-log}. The resulting
trace repeatedly halves the current value and checks whether it is equal to
one, or odd. In either of these cases the trace is left via a guard failure.

\begin{figure}
    \input{figures/unopt-log.tex}
    \caption{Unoptimized trace. Numbers on the right represent the line numbers in the original program. Lines marked with -1 represent lines added by JIT}
    \label{fig:unopt-trace}
\end{figure}

>>>>

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. There
are, besides the Python interpreter, implementations of Prolog, Javascript, R,
Smalltalk among other that are written in RPython at different levels of
completeness.

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as (if needed) a garbage collector
and based on hints provided by the author a just-in-time compiler.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing is a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

@@

Figure~\ref{fig:example} shows an example RPython function that checks
whether a number reduces to 1 with less than 100 steps of the Collatz process.\footnote{\url{http://en.wikipedia.org/wiki/Collatz_conjecture}}
It uses an \lstinline{Even} and an \lstinline{Odd} class to box the numbers, to
make the example more interesting. If the loop in \lstinline{check_reduces} is
traced when \lstinline{a} is a multiple of four, the unoptimized
trace looks like in Figure~\ref{fig:unopt-trace}. The line numbers in the trace
correspond to the line numbers in Figure~\ref{fig:trace-log}. The resulting
trace repeatedly halves the current value and checks whether it is equal to
one, or odd. In either of these cases the trace is left via a guard failure.

\begin{figure}
    \input{figures/unopt-log.tex}
    \caption{Unoptimized trace, the line numbers in the trace correspond to the line numbers in Figure~\ref{fig:trace-log}.}
    \label{fig:unopt-trace}
\end{figure}
